A Heuristic Approach towards Drawings of Graphs with High Crossing Resolution

### 2018-06-10 paper/paper.tex
paper: minor typos + email

\institute{
Wilhelm-Schickhard-Institut f\"ur Informatik, Universit\"at T\"ubingen, Germany\\
\texttt{\{bekos,foersth,mk\}@informatik.uni-tuebingen.de}\\
\texttt{\{christian-marius.geckeler,jan-lukas.hollaender,amadaeus.spallek,jan.splett\} @student.uni-tuebingen.de}
}

@@

Formally, the value of the minimum angle formed by any two crossing edges in a drawing is refered to as its \emph{crossing resolution}; the crossing resolution of a graph is defined as the maximum crossing resolution over all its drawings. Clearly, the crossing resolution of a non-planar graph is at most $90^\circ$, while a graph that admits a drawing with crossing resolution $90^\circ$ is called \emph{right-angle-crossing} graph or \emph{RAC} graph, for short; see Figure~\ref{fig:examples}. For these graphs, several results, mostly of theoretical nature, are known (refer to Section~\ref{sec:relatedwork} for a short overview). Notably, RAC graph are sparse (they contain at most $4n-10$ edges~\cite{DBLP:journals/tcs/DidimoEL11}, where $n$ denotes the number of  vertices), while deciding whether a graph is RAC is NP-hard~\cite{DBLP:journals/jgaa/ArgyriouBS12}.

@@

An immediate observation emerging from the above overview is that the~focus has been primarily on theoretical aspects of the problem. Most of the approaches that could be useful in practise are based on force-directed techniques~\cite{DBLP:books/ph/BattistaETT99,DBLP:journals/congnum/Eades84}. COWA is a system that supports conceptual web site traffic analysis~\cite{DBLP:conf/apvis/DidimoLR10}; its algorithmic core is a force-directed heuristic to compute simultaneous embeddings of two non-planar graphs with high crossing resolution. 
%
Didimo et al.~\cite{DBLP:conf/gd/DidimoLR10} describe topology-driven force-directed heuristics to achieve good trade-offs in terms of number of edge crossings, crossing resolution, and geodesic edge tendency; the obtained drawings, however, are not straight-line. 
%
For straight-line drawings, Nguyen et al.~\cite{DBLP:conf/gd/NguyenEHH10} suggest a quadratic-program to increase the crossing angles of circular drawings. 
%
Of more general scope are the already mentioned force-directed algorithms of Argyriou et al.\ and Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}.

@@

Our second approach is based on parameters $\rho$, $\delta_{min}$ and $\delta_{max}$ of the algorithm. Our idea was that if the algorithm gets trapped to a locally optimal solution, then a ``drastical'' or ``sharp'' move may help to escape. We turned this idea into an algorithmic implementation as follows: 
%
\begin{inparaenum}[(i)]
\item if during the last $\zeta$ iterations the crossing resolution has not been improved, we double the values of $\rho$, $\delta_{min}$ and $\delta_{max}$, and the algorithm is executed with these values for $\zeta'$ iterations;
\item afterwards, $\rho$, $\delta_{min}$ and $\delta_{max}$ switch back to this initial value.
\end{inparaenum}
%
Of course, this approach may lead to drawings with larger area, but this is expected, as it turns out that drawings with high crossing resolution may require large area~\cite{DBLP:journals/jgaa/AngeliniCDFBKS11,DBLP:journals/tcs/BrandenburgDEKL16}.

@@

% ==================================================================
\myparagraph{Total resolution} 
% ==================================================================
%
The notion of the total resolution of a drawing was introduced relatively recently with aim of ``balancing'' the measures of the crossing and of the angular resolution of a drawing~\cite{DBLP:journals/cj/ArgyriouBS13}. Formally, it is defined as the minimum of these two measures. It was not difficult to adjust our algorithm to yield drawings with high total resolution by simply taking into account also the angular resolution of the drawing. In particular, if the total resolution of the drawing is defined by its angular resolution, then the way we compute the critical vertices of this drawing has to change; the critical vertices must be the endvertices of the pairs of edges that define the angular resolution. Also, at each iteration of our algorithm we have to ensure that the total resolution does not decrease. We do so by rejecting candidate positions, which yield a reduced total~resolution.

@@

Regarding the crossing resulotion, we can observe that with increasing grid size, we could achieve better crossing resolution; see Fig.~\ref{fig:crossingResolutionGrid}. More precisely, the grid of size $10^2 \times 10^2$ was too restrictive for the vast majority of graphs which often resulted in the initial layout to be accepted as the best solution for larger graphs. Starting from grid size $10^3 \times 10^3$, we could observe, for small graphs, the performance was close to the performance of the unrestricted version of our algorithm (double line in Fig.~\ref{fig:crossingResolutionGrid}). For grid size $10^3 \times 10^3$, the performance declined for the larger graphs in the test set and we again could observe that for some of them the algorithm was not able to improve the initial layout. For grid size $10^4 \times 10^4$, the algorithm performed only about $10^\circ$ worse than the unrestricted version of our algorithm which could only slightly be improved by increasing the grid size to $10^6 \times 10^6$.

>>>>

\institute{
Wilhelm-Schickhard-Institut f\"ur Informatik, Universit\"at T\"ubingen, Germany\\
\texttt{\{bekos,foersth,geckeler,mk\}@informatik.uni-tuebingen.de}\\
\texttt{\{jan-lukas.hollaender,amadaeus.spallek,jan.splett\} @student.uni-tuebingen.de}
}

@@

Formally, the value of the minimum angle formed by any two crossing edges in a drawing is referred to as its \emph{crossing resolution}; the crossing resolution of a graph is defined as the maximum crossing resolution over all its drawings. Clearly, the crossing resolution of a non-planar graph is at most $90^\circ$, while a graph that admits a drawing with crossing resolution $90^\circ$ is called \emph{right-angle-crossing} graph or \emph{RAC} graph, for short; see Figure~\ref{fig:examples}. For these graphs, several results, mostly of theoretical nature, are known (refer to Section~\ref{sec:relatedwork} for a short overview). Notably, RAC graph are sparse (they contain at most $4n-10$ edges~\cite{DBLP:journals/tcs/DidimoEL11}, where $n$ denotes the number of  vertices), while deciding whether a graph is RAC is NP-hard~\cite{DBLP:journals/jgaa/ArgyriouBS12}.

@@

An immediate observation emerging from the above overview is that the~focus has been primarily on theoretical aspects of the problem. Most of the approaches that could be useful in practice are based on force-directed techniques~\cite{DBLP:books/ph/BattistaETT99,DBLP:journals/congnum/Eades84}. COWA is a system that supports conceptual web site traffic analysis~\cite{DBLP:conf/apvis/DidimoLR10}; its algorithmic core is a force-directed heuristic to compute simultaneous embeddings of two non-planar graphs with high crossing resolution.
%
Didimo et al.~\cite{DBLP:conf/gd/DidimoLR10} describe topology-driven force-directed heuristics to achieve good trade-offs in terms of number of edge crossings, crossing resolution, and geodesic edge tendency; the obtained drawings, however, are not straight-line. 
%
For straight-line drawings, Nguyen et al.~\cite{DBLP:conf/gd/NguyenEHH10} suggest a quadratic-program to increase the crossing angles of circular drawings. 
%
Of more general scope are the already mentioned force-directed algorithms of Argyriou et al.\ and Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}.

@@

Our second approach is based on parameters $\rho$, $\delta_{min}$ and $\delta_{max}$ of the algorithm. Our idea was that if the algorithm gets trapped to a locally optimal solution, then a ``drastic'' or ``sharp'' move may help to escape. We turned this idea into an algorithmic implementation as follows:
%
\begin{inparaenum}[(i)]
\item if during the last $\zeta$ iterations the crossing resolution has not been improved, we double the values of $\rho$, $\delta_{min}$ and $\delta_{max}$, and the algorithm is executed with these values for $\zeta'$ iterations;
\item afterwards, $\rho$, $\delta_{min}$ and $\delta_{max}$ switch back to this initial value.
\end{inparaenum}
%
Of course, this approach may lead to drawings with larger area, but this is expected, as it turns out that drawings with high crossing resolution may require large area~\cite{DBLP:journals/jgaa/AngeliniCDFBKS11,DBLP:journals/tcs/BrandenburgDEKL16}.

@@

% ==================================================================
\myparagraph{Total resolution} 
% ==================================================================
%
The notion of the total resolution of a drawing was introduced relatively recently with aim of ``balancing'' the measures of the crossing and of the angular resolution of a drawing~\cite{DBLP:journals/cj/ArgyriouBS13}. Formally, it is defined as the minimum of these two measures. It was not difficult to adjust our algorithm to yield drawings with high total resolution by simply taking into account also the angular resolution of the drawing. In particular, if the total resolution of the drawing is defined by its angular resolution, then the way we compute the critical vertices of this drawing has to change; the critical vertices must be the endvertices of the pairs of edges that define the angular resolution. Also, at each iteration of our algorithm we have to ensure that the total resolution does not decrease. We do so by rejecting candidate positions which yield a reduced total~resolution.

@@

Regarding the crossing resolution, we can observe that with increasing grid size, we could achieve better crossing resolution; see Fig.~\ref{fig:crossingResolutionGrid}. More precisely, the grid of size $10^2 \times 10^2$ was too restrictive for the vast majority of graphs which often resulted in the initial layout to be accepted as the best solution for larger graphs. Starting from grid size $10^3 \times 10^3$, we could observe, for small graphs, the performance was close to the performance of the unrestricted version of our algorithm (double line in Fig.~\ref{fig:crossingResolutionGrid}). For grid size $10^3 \times 10^3$, the performance declined for the larger graphs in the test set and we again could observe that for some of them the algorithm was not able to improve the initial layout. For grid size $10^4 \times 10^4$, the algorithm performed only about $10^\circ$ worse than the unrestricted version of our algorithm which could only slightly be improved by increasing the grid size to $10^6 \times 10^6$.
